---
title: "ENVS 193 DS Final"
author: "Evelyn Tsang"
date: "6/11/2025"
format: 
  html: 
    toc: true
    toc_float: true
---
# Part 1. Set up tasks

### Link to GitHub Repository: <https://github.com/ewtsang/ENVS-193DS_homework-03.git>
*change*

## Packages and Data:

```{r setup, message=FALSE}
#loading in libraries
library(tidyverse)
library(here)
library(gt)
library(janitor)
library(readxl)
library(dplyr)
library(ggplot2)
library(magick)
```

## Problem 1. Research Writing

### a. Transparent statistical methods

### b. More information needed

### c. Suggestions for rewriting

## Problem 2. Data visualization

### a. Cleaning and summarizing
```{r}
#create object called sst_clean from sst
```

### b. Visualize the data

## Problem 3. Data analysis

### a. Response variable

### b. Purpose of study

### c. Difference in "seasons"

### d. Table of models

### e. Run the models

### f. Check the diagnostics


### g. Select the best model

### h. Visualize the model preductions

### i. Write a caption for your figure


### j. Calculate model predictions

### k. Interpret your results

## Problem 4. Affective and exploratory visualizations

### a. Comparing visualizations

### b. Sharing you affective visualization
