---
title: "ENVS 193 DS Final"
author: "Evelyn Tsang"
date: "6/11/2025"
format: 
  html: 
    toc: true
    toc_float: true
---
# Part 1. Set up tasks

### Link to GitHub Repository: <https://github.com/ewtsang/ENVS-193DS_homework-03.git>
*change*

## Packages and Data:

```{r setup, message=FALSE}
#loading in libraries
library(tidyverse)
library(here)
library(gt)
library(janitor)
library(readxl)
library(dplyr)
library(ggplot2)
library(magick)
```

```{r}
#read in SB LTER data for this assignment (problem 2)
sst <-read.csv(here("data", "SST_update2023.csv"))
```

## Problem 1. Research Writing

### a. Transparent statistical methods

### b. More information needed

### c. Suggestions for rewriting

## Problem 2. Data visualization

"This data package contains sea surface temperature (SST) data in the Santa Barbara Channel area. Data was obtained from the NOAA National Centers for Environmental Information (NCEI) at 0.25° resolution for the time between 1982 and 2023. This Daily Optimum Interpolation Sea Surface Temperature (OISST) Analysis (Version 2.1) derived its data from satellite (Advanced Very High Resolution Radiometer (AVHRR)) and in situ platforms (i.e., ships and buoys) and yielded 18 gird points within the Santa Barbara Channel."

### a. Cleaning and summarizing
```{r}
#create object called sst_clean from sst
sst_clean <- sst |> 
  mutate(date = ymd(date),
         year = year(date), #converting to date, making sure in numeric values
         month = month(date, label = TRUE)) |> #abbreviated month "Jan", "Feb", etc.
  filter(year %in% c("2018", "2019", "2020", "2021", "2022", "2023")) |> #only use data from 2018 to 2023
  mutate(month = factor(month, levels = month.abb, ordered = TRUE)) |>  #ensure month is an ordered factor (Jan to Dec)
#need a data frame with abbreviated month, year, and mean_monthly_sst
  group_by(year, month) |> 
  summarize(mean_monthly_sst = round(mean(temp, na.rm = TRUE), 1), .groups = "drop")
#round temperatures to one decimal point
#drop duplicates

sst_clean$year <- factor(sst_clean$year, levels = c("2018", "2019", "2020", "2021", "2022", "2023"))
#display 5 rows from sst_clean
slice_sample(sst_clean, n = 5)
#structure from sst_clean
str(sst_clean)
```

### b. Visualize the data
```{r}
ggplot(data = sst_clean, #making visualization using cleaned sst data 
        aes(x = month, y = mean_monthly_sst, color = year, group = year)) +
  scale_color_manual(
    values = colorRampPalette(c("lightblue","darkblue"))(6)
  ) +
  geom_line(
    aes(group = year)) +
  geom_point(
    aes(), 
  #each individual point is sorted by year 
     size = 2) +
  #labeling the title and axis of the plot
labs(title = "Mean monthly surface temperature (C°) by year, 2018 to 2023",
       x = "Month",
       y = "Mean monthly sea surface temperature(C°)") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA), #adding a panel border
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"), #adding axis ticks
    legend.position = c(0.2, 0.8),
  )  #remove background gridlines and legend box
```


## Problem 3. Data analysis
Working with the nest box occupancy dataset from Stojanovic, D., Owens, G., Young, C.M., Alves, F. and Heinsohn, R. 2021. “Do nest boxes breed the target species or its competitors? A case study of a critically endangered bird.” Restoration Ecology. DOI: 10.1111/rec.13319

### a. Response variable

### b. Purpose of study

### c. Difference in "seasons"

### d. Table of models

### e. Run the models

### f. Check the diagnostics


### g. Select the best model

### h. Visualize the model preductions

### i. Write a caption for your figure


### j. Calculate model predictions

### k. Interpret your results

## Problem 4. Affective and exploratory visualizations

### a. Comparing visualizations

### b. Sharing you affective visualization
