---
title: "ENVS 193 DS Final"
author: "Evelyn Tsang"
date: "6/11/2025"
format: 
  html: 
    toc: true
    toc_float: true
---
# Part 1. Set up tasks

### Link to GitHub Repository: <https://github.com/ewtsang/ENVS-193DS_homework-03.git>
*change*

## Packages and Data:

```{r setup, message=FALSE}
#loading in libraries
library(tidyverse)
library(here)
library(gt)
library(janitor)
library(readxl)
library(dplyr)
library(ggplot2)
library(magick)
```

```{r}
#read in SB LTER data for this assignment (problem 2)
sst <-read.csv(here("data", "SST_update2023.csv"))
```

## Problem 1. Research Writing

### a. Transparent statistical methods

### b. More information needed

### c. Suggestions for rewriting

## Problem 2. Data visualization

"This data package contains sea surface temperature (SST) data in the Santa Barbara Channel area. Data was obtained from the NOAA National Centers for Environmental Information (NCEI) at 0.25° resolution for the time between 1982 and 2023. This Daily Optimum Interpolation Sea Surface Temperature (OISST) Analysis (Version 2.1) derived its data from satellite (Advanced Very High Resolution Radiometer (AVHRR)) and in situ platforms (i.e., ships and buoys) and yielded 18 gird points within the Santa Barbara Channel."

### a. Cleaning and summarizing
```{r}
#create object called sst_clean from sst
sst_clean <- sst |> 
  mutate(date = ymd(date),
         year = year(date), #converting to date
         month = month(date, label = TRUE)) |> #abbreviated month "Jan", "Feb", etc.
  filter(year %in% 2018:2023) |> #only use data from 2018 to 2023
  group_by(year, month) |> 
  summarize(mean_monthly_sst = mean(temp, na.rm = TRUE), .groups = "drop")
#need a data frame with abbreviated month, year, and mean_monthly_sst

#display 5 rows from sst_clean
slice_sample(sst_clean, n = 5)
#structure from sst_clean
str(sst_clean)
```

### b. Visualize the data

## Problem 3. Data analysis

### a. Response variable

### b. Purpose of study

### c. Difference in "seasons"

### d. Table of models

### e. Run the models

### f. Check the diagnostics


### g. Select the best model

### h. Visualize the model preductions

### i. Write a caption for your figure


### j. Calculate model predictions

### k. Interpret your results

## Problem 4. Affective and exploratory visualizations

### a. Comparing visualizations

### b. Sharing you affective visualization
