---
title: "ENVS 193 DS Final"
author: "Evelyn Tsang"
date: "6/11/2025"
format: 
  html: 
    toc: true
    toc_float: true
---
#Set up tasks

## Link to GitHub Repository: <https://github.com/ewtsang/ENVS-193DS_spring-2025_final>

## Packages and Data:

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
#loading in libraries
library(tidyverse) #general use
library(here) #organization
library(ggeffects) #model predictions
library(janitor) #cleaning data frames
library(readxl) #reading excel sheets
library(dplyr) #data manipulation
library(ggplot2) #creating figures
library(magick) #image processing
library(DHARMa) #regression modeling
library(MuMIn) #model selection
```

```{r reading-data}
#read in SB LTER data for this assignment (problem 2)
sst <-read.csv(here("data", "SST_update2023.csv"))

#read in Swift Parrot data for this assignment (problem 3)
sp_data <-read.csv(here("data", "occdist.csv"))
```

## Problem 1. Research Writing

### a. Transparent statistical methods
In part one, the coworker used a correlation test to assess that there is no relationship between distance from headwater (km) and annual total nitrogen load (kg year-1)

In part two, the coworker used a one-way ANOVA test to see whether average nitrogen load differs across multiple source types (urban land, atmospheric deposition, fertilizer, wastewater treatment, and grasslands)

### b. More information needed
What should also be included are (1) Effect size/dist for both analyses. For example, in part 1, was correlation positive? negative? how strong was it? A p-value by itself did not indicate the strength or nature of the relationship. (2) Post-hoc comparison for part 2 is needed. Since there are multiple nitrogen sources, a significant ANOVA should be followed by pairwise comparisons (Tukey's HSD) to determine difference in sources. This would give better context for interpretation of which alnd uses are contributing the most to nitrogen.

### c. Suggestions for rewriting
Sites father from the river headwaters were associated with higher total annual nitrogen loads. We rejected the null hypotheses that there is no correlation between distance from headwater (km) and annual total nitrogen load (kg year-1). Using Pearson's correlation, r = correlation coefficient, p = 0.03, a = 0.05

Nitrogen load varied significantly across different source types, indicating there is a key driver of elevated nitrogen levels to the system. We reject the null hypotheses that there is no difference in average nitrogen load between sources (urban land, atmospheric deposition, fertilizer, wastewater treatment, and grasslands) Using one way ANOVA, f=test statistic, df = degrees of freedom, p = 0.02, a = 0.05).A follow p post-hoc Tukey's HSD test determined that ___ source (mean difference = average kg/year, p = p value) had significantly higher nitrogen loads than ___source (mean difference = average kg/year, p = p-value) no significant differences were found between other sources.

## Problem 2. Data visualization

"This data package contains sea surface temperature (SST) data in the Santa Barbara Channel area. Data was obtained from the NOAA National Centers for Environmental Information (NCEI) at 0.25° resolution for the time between 1982 and 2023. This Daily Optimum Interpolation Sea Surface Temperature (OISST) Analysis (Version 2.1) derived its data from satellite (Advanced Very High Resolution Radiometer (AVHRR)) and in situ platforms (i.e., ships and buoys) and yielded 18 gird points within the Santa Barbara Channel."

### a. Cleaning and summarizing
```{r data-cleaning}
#create object called sst_clean from sst
sst_clean <- sst |> 
  mutate(date = ymd(date),
         year = year(date), #converting to date, making sure in numeric values
         month = month(date, label = TRUE)) |> #abbreviated month "Jan", "Feb", etc.
  filter(year %in% c("2018", "2019", "2020", "2021", "2022", "2023")) |> #only use data from 2018 to 2023
  mutate(month = factor(month, levels = month.abb, ordered = TRUE)) |>  #ensure month is an ordered factor (Jan to Dec)
#need a data frame with abbreviated month, year, and mean_monthly_sst
  group_by(year, month) |> 
  summarize(mean_monthly_sst = round(mean(temp, na.rm = TRUE), 1), .groups = "drop")
#round temperatures to one decimal point
#drop duplicates

sst_clean$year <- factor(sst_clean$year, levels = c("2018", "2019", "2020", "2021", "2022", "2023"))
#display 5 rows from sst_clean, in levels ("2018", "2019", "2020", "2021", "2022", and "2023"
slice_sample(sst_clean, n = 5)
#structure from sst_clean
str(sst_clean)
```

### b. Visualize the data
*how to add20 tick to yaxis*
```{r sst-visualization}
ggplot(data = sst_clean, #making visualization using cleaned sst data 
        aes(x = month, y = mean_monthly_sst, color = year, group = year)) + 
  #setting the axis, comparing month (x axis) and mean monthly sea surface temperature (y axis)
  scale_color_manual(
    values = colorRampPalette(c("lightblue","darkblue"))(6)
  ) + 
  #color gradient representing years
  geom_line(
    aes(group = year)) + 
  #group by
  geom_point(
    aes(), 
  #each individual point is sorted by year 
     size = 2) +
  #labeling the title and axis of the plot
labs(title = "Mean monthly surface temperature (C°) by year, 2018 to 2023",
       x = "Month",
       y = "Mean monthly sea surface temperature(C°)") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA), #adding a panel border
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"), #adding axis ticks
    legend.position = c(0.2, 0.7),
  )  #remove background gridlines and legend box
```


## Problem 3. Data analysis

Working with the nest box occupancy dataset from Stojanovic, D., Owens, G., Young, C.M., Alves, F. and Heinsohn, R. 2021. “Do nest boxes breed the target species or its competitors? A case study of a critically endangered bird.” Restoration Ecology. DOI: 10.1111/rec.13319

### a. Response variable
In this dataset, a 1 or 0 marks nest box occupancy for the columns of either Swift Parrot, Common Starling, Tree Martins, or if the the nest box was empty. A 1 indicates occupancy and 0 indicates vacancy.

### b. Purpose of study
The study investigates how nest box occupancy by Swift Parrots and competing species (Tree Martins and Common Starlings) is influenced by distance from the forest edge and time since deployment. The researchers aim to evaluate whether permanent nest boxes benefit the endangered Swift Parrot or unintentionally support competitors.

### c. Difference in "seasons"
In 2016, the nest boxes were newly deployed, while in 2019 they were established and had been in place for 3 years. The years represent different breeding seasons and help assess changes in box occupancy over time.

### d. Table of models

| Model number | Edge Distance | Season          | Description                 |  
|:------------:|:-------------:|:---------------:|:----------------------------|  
| 0            |               |                 | no predictors (null model)  |
| 1            |       X       |                 | edge distance aspredictor   | 
| 2            |               |       X         | season as predictor         |   
| 3            |       X       |       X         | all predictors (full model) |    

### e. Run the models
```{r, results="hide"}
sp_data_clean <- clean_names(sp_data)  

model10 <- lm(sp ~ 1, data = sp_data_clean) #nopredictors

model11 <- lm(sp ~ edge_distance, data = sp_data_clean) #edge distance

model12<- lm(sp ~ season, data = sp_data_clean) #season

model13 <- lm(sp ~ edge_distance+  season, data = sp_data_clean) #both predictors
```

### f. Check the diagnostics

#### Model 0, Null model
```{r}
sim0 <- simulateResiduals(model10)
par(mfrow = c(2,2))
plot(sim0)
```

#### Model 1, Distance as predictor
```{r}
sim1 <- simulateResiduals(model11)
par(mfrow = c(2,2))
plot(sim1)

```

#### Model 2, Season as predictor
```{r}
sim2 <- simulateResiduals(model12)
par(mfrow = c(2,2))
plot(sim2)

```

#### Model 3, Both season and distance as predictor
```{r}
sim3 <- simulateResiduals(model13)
par(mfrow = c(2,2))
plot(sim3)

```

### g. Select the best model
```{r model-fitting}
AICc(model10, model11, model12, model13) |> 
  arrange(AICc)
```

The best model to use would be **Model 3**. 

### h. Visualize the model predictions

```{r}
library(ggeffects)
sp_data_clean <- sp_data_clean |>
  mutate(season = as.factor(season))

# Generate predicted values for combinations of edge_distance and season
preds <- ggpredict(model13, terms = c("edge_distance [all]", "season"))

# Plot with ribbons, predictions, and jittered observed data
ggplot(preds, aes(x = x, y = predicted, color = as.factor(group), fill = as.factor(group))) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
  geom_jitter(data = sp_data_clean, aes(x = edge_distance, y = sp, color = season), 
              inherit.aes = FALSE, width = 20, height = 0.05, alpha = 0.3) +
  labs(title = "Predicted Swift Parrot Nest Box Occupancy by Distance and Season",
       x = "Distance from Forest Edge (m)",
       y = "Probability of Swift Parrot Occupancy",
       color = "Season",
       fill = "Season") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(color = "black", fill = NA),
        axis.line = element_line(color = "black"),
        axis.ticks = element_line(color = "black"),
        legend.position = "right") +
scale_color_manual(values = c("2016" = "orange", "2019" = "yellowgreen")) +
  scale_fill_manual(values = c("2016" = "orange", "2019" = "yellowgreen")) 

```

### i. Write a caption for your figure
*Figure 1.* Predicted probability of Swift Parrot nest box occupancy based on distance from forest edge and on season (2016 vs. 2019). Lines represent model predictions from the all predictors model (occupancy ~ edge_distance + season), and shaded areas are 95% confidence intervals. Points represent raw occupancy data, jittered for visibility. Data source: Stojanovic, Dejan et al. (2021). Do nest boxes breed the target species or its competitors? A case study of a critically endangered bird [Dataset]. Dryad. https://doi.org/10.5061/dryad.83bk3j9sb

### j. Calculate model predictions
```{r}
pred <- ggpredict(model13, terms = c("edge_distance [0, 9000]", "season"))
print(pred)
```

### k. Interpret your results
Swift Parrots were less likely to use boxes farther from forest edge.
Occupancy probability was higher in 2016 (new boxes) than 2019.
Likely because permanent boxes were learned and overtaken by competitors (esp. starlings).
Management should prioritize temporary or seasonally-sealed box

## Problem 4. Affective and exploratory visualizations

Homework 3 Sketch
```{r}
image<- image_read("/Users/ewtsa/Documents/ENVS-193DS/GitHub/ENVS-193DS_homework-03/HW3Paper.jpg")
grid::grid.raster(image)
```

Homework 3 Draft
```{r}
image<- image_read("/Users/ewtsa/Documents/ENVS-193DS/GitHub/ENVS-193DS_homework-03/HW3Draft.jpg")
grid::grid.raster(image)
```

Final Rendition

```{r}
image<- image_read("/Users/ewtsa/Documents/ENVS-193DS/GitHub/ENVS-193DS_homework-03/HW3Draft.jpg")
grid::grid.raster(image)
```

### a. Comparing visualizations


### b. Sharing you affective visualization

